Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-17 doc/mc-paper.tex
Added lemma example and polished

Model counting is a challenging problem---more challenging that the
already NP-hard problem of Boolean satisfiability.  Several
tractable variants of Boolean satisfiability, including 2-SAT, become
intractable when the goal becomes model counting and not just
satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
model counters have developed that scale to very large formulas, as
witnessed by their progress in the annual model counting competitions.

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
becomes straightforward.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by Boolean operation And or Or.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal of the graph.
Kimmig, et al., present a very general framework, known as {\em
  algebraic model counting}~\cite{kimmig:jal:2017}, describing
properties of Boolean functions that can be efficiently computing from
its d-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations commonly used in knowledge
compilation, including d-DNNF\@.  We present the CRAT file format.
It captures both the structure of a POG and a checkable proof
that the POG is logically equivalent to a Boolean formula given in
conjunctive normal form (CNF).  A proof consists of a sequence of
clause addition and deletion steps, based on the extended resolution
proof system~\cite{Tseitin:1983}.  We present a set of conditions that
a CRAT file must satisfy to provide an equivalence proof.

@@

Figure~\ref{fig:chain} illustrates the certifying knowledge compilation
and (weighted) model counting tool chain that we have developed.  A
{\em proof generator} generates a CRAT description from a d-DNNF
graph generated by \dfour{}, a state-of-the-art knowledge compiler.
A {\em proof checker} then verifies that the combination of CNF and CRAT files
satisfy our rules, and a {\em ring evaluator} can compute
either a standard or weighted model count from the POG\@ representation.
We evaluate the capabilities of this tool chain using
benchmark formulas from the 2022 standard and weighted model
competitions.  Our tools are able to handle all but the largest d-DNNF
graphs generated by \dfour{}.
As the dashed box in Figure~\ref{fig:chain} indicates, this tool chain
shifts the need for trustworthy code away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trustworthy---its errors will also be caught by the proof checker.

@@

Having a SAT solver generate a proof of unsatisfiability has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently (and, in some cases, formally) checked
\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress has also
been made on extending these proof systems to quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound and complete.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  Our tool chain is not directly
comparable to theirs---they certify the output of a standard modeling
program, while we certify the output of a knowledge compiler and then
produce a certified standard or weighted model count.  Their tool
chain can be applied to some programs that are pure model counters,
while ours can be applied to pure knowledge compilers.  Nonetheless,
we compare the performance of our tool chain to theirs as part of our
experimental evaluation in Section~\ref{sect:experimental}.  Our
experience with developing a formally verified proof checker has shown
us that, even for the well understood framework of extended
resolution, it can be challenging to formulate the full set of requirements to guarantee soundness.

We know of know other work addressing the more general question of
certifying the output of a knowledge compiler.

@@

We propose a general framework for summarizing properties of Boolean
formulas inspired by the formulation of {\em algebraic model counting}.~\cite{kimmig:jal:2017}.

@@

  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum $\sum$ is computed using addition operation $\radd$, and product $\prod$ is computed using multiplication operation $\rmul$.
\label{def:labeling}
\end{definition}

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
{\Large \rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
{\Large \rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}

@@

The CRAT format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
The following are its key properties:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{\nodeu}$
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our tool chain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node.  Rather than representing every
d-DNNF node explicitly, our program performs simplifications to
eliminate Boolean constants.
Except for the degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

The performance of the proof generator for forward implication, both in its execution time and
the size of the proof generated, can be improved by two optimizations
described here.  A key feature is that they do not require any changes
to the proof framework---they build on the power of extended
resolution to create new logical structures within the proof.  They
involve declaring new product nodes to encode both products of literals.
These nodes are not part of the POG
representation of the formula; they serve only to enable the forward
implication proof.  Here we only summarize the two optimizations.
More details are presented in Appendix~\ref{app:optimizations}.

@@

{\em Lemmas:} Our formulation of $\validate$ requires each call at a
node $\nodeu$ to recursively validate all of its children.  This
effectively expands the graph into a tree, potentially requiring an
exponential number of recursive steps.  Instead, for each node
$\nodeu$ having $\indegree(u) > 1$, the program can define and
generate the proof of a lemma for $\nodeu$ when it is first reached by
a call to $\validate$ and then apply this lemma for this and
subsequent calls.  The lemma states that the POG with root node
$\nodeu$ satisfies forward implication for a formula
$\gamma_{\nodeu}$, where some of the clauses in this formula can be
input clauses from $\inputformula$, but others can be simplified
versions of input clauses.
The key idea is to introduce product nodes to encode (via DeMorgan's
Laws) the simplified clauses and have these serve as lemma arguments.

@@

\section{Experimental Evaluation}
\label{sect:experimental}
We have implemented versions of the three programs that, in addition to the \dfour{} knowledge compiler,  form the
toolchain illustrated in Figure~\ref{fig:chain}.  One version,
including a proof generator, proof checker, and ring evaluator for
standard and weighted model counting, is designed to be fully
scalable.  The other version, consisting of a proof checker and
ring evaluator has been formally verified within the \lean{} proof
system.

Our long-term goal is to have a toolchain suitable for use in a
knowledge compilation competition, along the lines of the annual
competitions for other automated reasoning tools.  Our experiments
consider the benchmark problems and computing capabilities that would
arise in such a competition and evaluate whether our implementations
could handle these problems.  As with the original proof-generating
SAT competitions, we assume that the proofs at a competition could be
checked with the unverified checker.  Having a verified checker
that works on some subset of the benchmarks would provide an
additional degree of certainty.

@@

The proof generator can optionally be instructed to generate a {\em
one-sided} proof, providing only the reverse-implication portion of the proof via
input clause deletion.  This can provide useful information---any
assignment that is a model for the compiled representation
must also be a model for the input formula---even when generating a
full validation is impractical.

Our ring evaluator can perform both standard and weighted model
counting.  It performs arithmetic over a subset of the rational
numbers we term $\drational$, consisting of numbers of the form
$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
with $a$ implemented to have arbitrary range.
Allowing scaling by powers of 2 enables the density computation
and rescaling required for standard model counting.  Allowing scaling
by powers of both 2 and 5 enables exact decimal arithmetic, handling
the weights used in the weighted model counting competitions.

@@

@@

\subsection{Lemma Example}
Observe that the forward implication proof shown in Figure~\ref{fig:eg4:all}  must ``visit'' nodes
$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).

@@

Our default configuration for the proof generator used the structural
method for the forward implication proof, with the two optimizations
of literal grouping and lemmas.
Running with a time limit of 4000 seconds\footnote{All measured times
in this document are actual elapsed times, not CPU times.}, \dfour{}
was able to generated d-DNNF representations for 124 of these, and it timed out on the
rest.  Our proof generator was able to convert all of the generated
d-DNNF graphs into POGs and determine how many defining clauses they
would generate.  A POG operation with $k$ arguments requires $k+1$
defining clauses, and so the total number of defining clauses is equal
to the sum of the number of nonterminal nodes plus the sum of number
of children for each of these nodes.  The number of defining clauses
ranged from 304 to 2,761,457,765 with a median of 848,784.

@@

For each of the 124 d-DNNF graphs generated by \dfour{}, we attempted
to use our proof generator and unverified checker, limiting proof
generation to 10,000 seconds.  We also attempted to generate and check
one-sided proofs for 124 graphs, again with a time limit of 10,000
seconds for proof generation.

@@

Relative to the runtime for \dfour{}, two problems ran faster
with the generator and checker.  These were for problems having small numbers of
models (relative to the number of variables), and so most of the time
spent by both \dfour{} and our proof generator was in running a CDCL
solver to search the very sparse solution space.  Our solver,
\cadical{} generally outperforms the miniSAT-based CDCL solver used by
\dfour{}.  At the other extreme, one problem that required only 19
seconds for \dfour{} required 8657 seconds to generate a proof and 45
seconds to check it.  It is unclear why the proof generation required
so much time.  Overall the ratios between the combined generation plus
checking times versus the time for \dfour{} had a harmonic mean of
5.5.

Of the 16 benchmarks that could not be fully validated, one had so
many defining clauses that it overflowed the 32-bit signed integers
our programs used for clause identifiers.  Another exited due to the
virtual memory limit imposed by the operating system, and the
other 14 timed out.  Of the 15 that did not overflow the clause
counter, we were able to complete a one-sided verification for 9, but
the other 6 timed out.

@@

Figure~\ref{fig:monolithic:clauses} shows the total numbers of clauses
for the two approaches for the 70 cases where both approaches completed within 1000 seconds.
As can be seen, the monolithic approach consistently produces
shorter proofs, perhaps due to the benefits of the proof trimming by
\dtrim{}.

@@

Overall, it is clear that these two optimizations are worthwhile, and
even critical to success for some benchmarks.

>>>>

Model counting is a challenging problem---more challenging that the
already NP-hard problem of Boolean satisfiability.  Several
tractable variants of Boolean satisfiability, including 2-SAT, become
intractable when the goal becomes model counting and not just
satisfiability \cite{valiant:siam:1979}.  Nonetheless, a number of
model counters have developed that scale to very large formulas, as
witnessed by the progress in recent model counting competitions.

One approach to model counting, known as {\em knowledge compilation},
transforms the formula into a structured form for which model counting
becomes straightforward.  For example, the {\em deterministic,
  decomposable, negation normal form} (d-DNNF) introduced by
Darwiche~\cite{darwiche:aaai:2002,darwiche:ecai:2004} represents a
Boolean formula as a directed acyclic graph, with terminal nodes
labeled by Boolean variables and their complements, and with each
nonterminal node labeled by Boolean operation And or Or.  Restrictions
are placed on the structure of the graph such that a count of the
models can be computed by a single bottom-up traversal.
Kimmig, et al., present a very general framework, known as {\em
  algebraic model counting}~\cite{kimmig:jal:2017}, describing
properties of Boolean functions that can be efficiently computing from
a d-DNNF representation.  These include standard and weighted model
counting, and much more.

@@

In this paper, we present \emph{Partitioned Operation Graphs} (POGs),
a generalization of the representations commonly used in knowledge
compilation, including d-DNNF\@.  We present the CRAT file format.
It captures both the structure of a POG and a checkable proof
that the POG is logically equivalent to a Boolean formula given in
conjunctive normal form (CNF).  A proof consists of a sequence of
clause addition and deletion steps, based on an extended resolution
proof system~\cite{Tseitin:1983}.  We present a set of conditions that
a CRAT file must satisfy to provide an equivalence proof.

@@

Figure~\ref{fig:chain} illustrates the certifying knowledge compilation
and (weighted) model counting tool chain that we have developed.  A
{\em proof generator} generates a CRAT description from a d-DNNF
graph generated by \dfour{}, a state-of-the-art knowledge compiler.
A {\em proof checker} then verifies that the combination of CNF and CRAT files
satisfy our rules, and a {\em ring evaluator} can compute
either a standard or weighted model count from the POG\@ representation.
We evaluate the capabilities of this tool chain using
benchmark formulas from the 2022 standard and weighted model
competitions.  Our tools are able to handle all but the largest d-DNNF
graphs generated by \dfour{}.
As the dashed box in Figure~\ref{fig:chain} indicates, this tool chain
shifts the need for trustworthy code away from the complex and highly
optimized knowledge compiler to a relatively simple checker and
evaluator.  Importantly, the proof generator need not be
trusted---its errors will also be caught by the proof checker.

@@

Having a SAT solver generate a proof of unsatisfiability has a long
tradition~\cite{ZhangMalik} and has become widely accepted due to the
formulation of clausal proof systems for which proofs can readily be
generated and efficiently (and, in some cases, formally) checked
\cite{cruz-cade-2017,RAT,Tan:2021,wetzler14_drattrim}.  Some progress has also
been made on extending these proof systems to other domains, including quantified Boolean
formulas~\cite{bryant:cade:2021,heule:JAR2014}.

Fichte, Hecher, and Roland~\cite{fichte:sat:2022} devised the MICE
proof framework for model counting programs.  Their proof rules are
based on the algorithmic steps commonly used by model counting
programs, and they present a series of lemmas and a theorem stating
that their checking framework is sound.  They also
present experimental results for cases where they modify existing
model counters to also generate proof traces, and where they convert
d-DNNF graphs into a checkable form.  Our tool chain is not directly
comparable to theirs---they certify the output of a standard modeling
program, while we certify the output of a knowledge compiler and then
produce a certified standard or weighted model count.  Their tool
chain can be applied to some programs that are pure model counters,
while ours can be applied to pure knowledge compilers.  Nonetheless,
we compare the performance of our tool chain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.  Our
experience with developing a formally verified proof checker has shown
us that, even for the well understood framework of extended
resolution, it can be challenging to formulate the full set of requirements to guarantee soundness.

We know of no other work addressing the question of
certifying the output of a knowledge compiler.

@@

We propose a general framework for summarizing properties of Boolean
formulas inspired by the formulation of {\em algebraic model counting}~\cite{kimmig:jal:2017}.

@@

  For Boolean formula $\phi$, and ring labeling $w$, the {\em ring evaluation problem} is to compute:
  \begin{eqnarray}
    \rep(\phi, w) & = & \sum_{\alpha \in \modelset(\phi)} \;\; \prod_{\lit \in \alpha} w(\ell) \label{eqn:rep}
  \end{eqnarray}
  In this equation, sum \scalebox{0.8}{$\sum$} is computed using addition operation $\radd$, and product \scalebox{0.8}{$\prod$} is computed using multiplication operation $\rmul$.
\label{def:labeling}
\end{definition}

@@

Performing ring evaluation on an arbitrary Boolean formula could be intractable, but it is straightforward for a formula with partitioned operations:
\begin{proposition}
\label{prop:ring:eval}
Ring evaluation with operations $\boolnot$, $\pand$, and $\por$ satisfies the following for any labeling $w$:
\begin{eqnarray}
%% V1
\textstyle
\rep(\boolnot \phi,\; w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Pand_{i=1,k} \phi_i,\; w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
\textstyle
\scalebox{1.2}{\rep}\left(\Por_{i=1,k} \phi_i,\; w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
%% V2
%\rep(\boolnot \phi, w) &=& \mulident - \rep(\phi, w) \label{eqn:ring:negation} \\
%\rep\left(\phi_1, w \pand \dots \pand \phi_k, w \right) &=& \prod_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\phi_1, w \por \dots \por \phi_k, w \right) &=& \sum_{i=1..k} \rep(\phi_i, w) \label{eqn:ring:sum}
%\rep\left(\Pand_{i=1,k} \phi_i, w \right) &=& \prod_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:product} \\
%\rep\left(\Por_{i=1,k} \phi_i, w\right) &=& \sum_{i=1,k} \rep(\phi_i, w) \label{eqn:ring:sum}
\end{eqnarray}
\end{proposition}

@@

The CRAT format draws its inspiration from the LRAT~\cite{lrat} and
QRAT~\cite{heule:JAR2014} formats for unquantified and quantified Boolean formulas, respectively.
Key properties include:
\begin{itemize}
  \item
  The file contains declarations of $\pand$ and $\por$ operations to describe the POG.
  Declaring a node $\nodeu$
implicitly adds an {\em extension} variable $u$ and a set of {\em defining} clauses $\theta_{\nodeu}$
  encoding the product or sum operation.
  This is the only means for adding extension variables to the proof.
\item Boolean negation is supported implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\item
  The file contains explicit clause addition steps.
  A clause can only be added if it is logically implied by the existing clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item
  The file contains explicit clause deletion steps.
  A clause can only be deleted if it is logically implied by the remaining clauses.
  A sequence of clause identifiers must be listed as {\em hints} providing a RUP verification of the implication.
\item The checker must track the dependency set for every input and
  extension variable.  For each $\pand$ operation, the checker must ensure that the dependency sets for its arguments are disjoint.
  The associated extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.
  Only binary $\por$ operations are allowed.  Generalizing to $k$ arguments would require
  $k\,(k-1)/2$ proofs of disjointness.
\end{itemize}

@@

%%Figure~\ref{fig:chain} illustrates one method of generating a CRAT
%%file from an input formula in conjunctive normal form.  Our tool chain
%%is based on the \dfour{} knowledge compiler~\cite{lagniez:ijcai:2017},
%%but it would also work for other tools that use a top-down search
%%strategy to generate a d-DNNF representation.
Converting a d-DNNF
graph into a POG is straightforward---each node in a d-DNNF graph can
be directly represented as a POG node.  Rather than representing every
d-DNNF node explicitly, our program performs simplifications to
eliminate Boolean constants.
Except in degenerate cases,
%where the formula is unsatisfiable or a tautology,
we can therefore assume
that the POG does not contain any constant nodes.
In addition, negation is only
applied to variables, and so we can merge the negation operations into the terminal nodes,
viewing the POG as consisting
of {\em literal} nodes corresponding to input variables and their negations, along with
{\em terminal} nodes, which can be further classified as {\em product} nodes and {\em sum} nodes.

@@

The performance of the structural proof generator for forward implication, both in its execution time and
the size of the proof generated, can be improved by two optimizations
described here.  A key feature is that they do not require any changes
to the proof framework---they build on the power of extended
resolution to create new logical structures within the proof.  They
involve declaring new product nodes to encode products of literals.
These nodes are not part of the POG
representation of the formula; they serve only to enable the forward
implication proof.  Here we only summarize the two optimizations.
More details are presented in Appendix~\ref{app:optimizations}.

@@

{\em Lemmas:} Our formulation of $\validate$ requires each call at a
node $\nodeu$ to recursively validate all of its children.  This
effectively expands the graph into a tree, potentially requiring an
exponential number of recursive steps.  Instead, for each node
$\nodeu$ having $\indegree(u) > 1$, the program can define and
generate the proof of a {\em lemma} for $\nodeu$ when it is first reached by
a call to $\validate$ and then apply this lemma for this and
subsequent calls.  The lemma states that the POG with root node
$\nodeu$ satisfies forward implication for a formula
$\gamma_{\nodeu}$, where some of the clauses in this formula can be
input clauses from $\inputformula$, but others can be simplified
versions of input clauses.
The key idea is to introduce product nodes to encode (via DeMorgan's
Laws) the simplified clauses and have these serve as lemma arguments.

@@

\section{Experimental Evaluation}
\label{sect:experimental}
We have implemented versions of the three programs that, along with the \dfour{} knowledge compiler,  form the
toolchain illustrated in Figure~\ref{fig:chain}.  One version,
including a proof generator, proof checker, and ring evaluator for
standard and weighted model counting, is designed to be fully
scalable.  The other version, consisting of a proof checker and
ring evaluator has been formally verified within the \lean{} proof
system.
Our long-term goal is to have a toolchain suitable for use in a
knowledge compiler competition, along the lines of the annual
competitions for other automated reasoning tools.  Our experiments
consider the benchmark problems and computing capabilities that would
arise in such a competition and evaluate whether our implementations
could handle these problems.  As with the original proof-generating
SAT competitions, we assume that the proofs at a competition could be
checked with the unverified checker.  Having a verified checker
that works on some subset of the benchmarks would provide an
additional degree of certainty.

@@

The proof generator can optionally be instructed to generate a {\em
one-sided} proof, providing only the reverse-implication portion of the proof via
input clause deletion.  This can provide useful information---any
assignment that is a model for the compiled representation
must also be a model for the input formula---even when 
full validation is impractical.

Our ring evaluator can perform both standard and weighted model
counting.  It performs arithmetic over a subset of the rationals
consisting of numbers of the form
$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
with $a$ implemented to have arbitrary range.
Allowing scaling by powers of 2 enables the density computation
and rescaling required for standard model counting.  Allowing scaling
by powers of both 2 and 5 enables exact decimal arithmetic, handling
the weights used in the weighted model counting competitions.

@@

Observe that the forward implication proof shown in Figure~\ref{fig:eg4:all}  must ``visit'' nodes
$\nodep_5$ and $\nodep_6$ twice, separately considering  assignments where $\assign(x_1) = 1$
(clauses 25 and 26) and where $\assign(x_1) = 1$ (clauses 30 and 31).  
Section~\ref{app:lemma:eg} shows how to use a lemma such that these nodes are each only visited once.

@@

\subsection{Lemma Example}
\label{app:lemma:eg}

\begin{figure}
(A) Additional nodes\\[1.0em]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{CRAT line} & Explanation \\
\midrule
25 & \texttt{p 11 -3 4} & \texttt{0} & $v_{11} = \obar{x}_3 \pand {x}_4$ \\
28 & \texttt{p 12  3 -4} & \texttt{0} & $v_{12} = {x}_3 \pand \obar{x}_4$ \\
\bottomrule
\end{tabular}
\\[1.0em]
(B) Implicit Clauses\\[1.2em]
\begin{tabular}{llll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clauses} & Explanation \\
\midrule
\texttt{25} & \texttt{11 3 -4} & \texttt{0} & Argument clause $\{x_3 ,\, \obar{x}_4\}$, activated by $\obar{v}_{11}$ \\
\texttt{26} & \texttt{-11 -3} & \texttt{0} & \\
\texttt{27} & \texttt{-11 4} & \texttt{0} & \\
\midrule
\texttt{28} & \texttt{12 -3 4} & \texttt{0} & Argument clause $\{\obar{x}_3,\,  {x}_4\}$, activated by $\obar{v}_{12}$ \\
\texttt{29} & \texttt{-12 3} & \texttt{0} & \\
\texttt{30} & \texttt{-12 -4} & \texttt{0} & \\
\bottomrule
\end{tabular}
\\[1.0em]
(C) CRAT Assertions\\[1.0em]
\begin{tabular}{llllll}
\toprule
\makebox[5mm]{ID} & \multicolumn{2}{l}{Clause} & \multicolumn{2}{l}{Hints} & Explanation \\
\midrule
\multicolumn{6}{l}{Lemma Proof} \\
\texttt{31} & \texttt{a 5 11 12 3} & \texttt{0} & \texttt{25 6} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply p_{5}$ \\
\texttt{32} & \texttt{a 6 11 12 -3} & \texttt{0} & \texttt{28 9} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land {x}_3 \imply p_{6}$ \\
\texttt{33} & \texttt{a 3 7 11 12} & \texttt{0} & \texttt{13 31} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \land \obar{x}_3 \imply s_{7}$ \\
\texttt{34} & \texttt{a 7 11 12} & \texttt{0} & \texttt{33 14 32} & \texttt{0} & $(\obar{v}_{11} \land \obar{v}_{12}) \imply s_{7}$ \\
\midrule
\multicolumn{6}{l}{Lemma Application \#1} \\
\texttt{35} & \texttt{a -11 1} & \texttt{0} & \texttt{26 27 3} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{11}$ \\
\texttt{36} & \texttt{a -12 1} & \texttt{0} & \texttt{29 30 4} & \texttt{0} & $\obar{x}_1 \imply \obar{v}_{12}$ \\
\texttt{37} & \texttt{a 7 1} & \texttt{0} & \texttt{35 36 34} & \texttt{0} & $\obar{x}_1 \imply s_7$ \\
\midrule
\texttt{38} & \texttt{a 8 1} & \texttt{0} & \texttt{37 15} & \texttt{0} & $\obar{x}_1 \imply p_8$ \\
\midrule
\multicolumn{6}{l}{Lemma Application \#2} \\
\texttt{39} & \texttt{a -11 -1} & \texttt{0} & \texttt{26 27 1} & \texttt{0} & ${x}_1 \imply \obar{v}_{11}$ \\
\texttt{40} & \texttt{a -12 -1} & \texttt{0} & \texttt{29 30 2} & \texttt{0} & ${x}_1 \imply \obar{v}_{12}$ \\
\texttt{41} & \texttt{a 7 -1} & \texttt{0} & \texttt{39 40 34} & \texttt{0} & ${x}_1 \imply s_7$ \\
\midrule
\texttt{42} & \texttt{a 9 -1} & \texttt{0} & \texttt{5 41 18} & \texttt{0} & ${x}_1 \imply p_9$ \\
\texttt{43} & \texttt{a 1 10} & \texttt{0} & \texttt{23 38} & \texttt{0} & $\obar{x}_1 \imply s_{10}$  \\
\texttt{44} & \texttt{a 10} & \texttt{0} & \texttt{43 24 42} & \texttt{0} & $s_{10}$ \\
\bottomrule
\end{tabular}
\caption{Example of lemma definition, proof, and application}
\label{fig:eg4:lemmas}
\end{figure}

Figure~\ref{fig:eg4:lemmas} shows an alternate forward implication
proof for the example of Figure~\ref{fig:eg4:all} using a lemma to
represent the shared node $\nodes_7$.  We can see that the POG with
this node as root encodes the Boolean formula $x_3 \leftrightarrow x_4$, having a CNF representation consisting of the clauses
$\{x_3 ,\, \obar{x}_4\}$ and $\{\obar{x}_3 ,\, {x}_4\}$.  The product node
declarations shown in Figure~\ref{fig:eg4:lemmas}(A) create synthetic
clauses 25 and 28 to encode these arguments with activating literals
$\obar{v}_{11}$ and $\obar{v}_{12}$, respectively.  Clauses 31--34
then provide a proof of the lemma, stating that any assignment
$\assign$ that activates these clauses will  assign $1$ to $s_7$.
Clauses 35 and 36 state that an assignment with $\assign(x_1) = 0$
will cause the first synthetic clause to activate due to input clause
3, and it will cause the second synthetic clause to activate due to
input clause 4.  From this, clause 37 can use the lemma to state that
assigning $0$ to $x_1$ will cause $s_7$ to evaluate to $1$.  Similarly,
clauses 39 and 40 serve to activate the synthetic clauses when
$\assign(x_1) = 1$, due to input clauses 1 and 2, and clause 41 then
uses the lemma to state that assigning $1$ to $x_1$ will cause $s_7$ to
evaluate to $1$.

In this example, adding the lemma increases the proof length, but that
is only because it is such a simple formula.

@@

Our default configuration for the proof generator used the structural
method for the forward implication proof, with the two optimizations
of literal grouping and lemmas.
Running with a time limit of 4000 seconds\footnote{All measured times
in this document are actual elapsed times, not CPU times.}, \dfour{}
was able to generate d-DNNF representations for 124 of these, and it timed out on the
rest.  Our proof generator was able to convert all of the generated
d-DNNF graphs into POGs and determine how many defining clauses they
would generate.  A POG operation with $k$ arguments requires $k+1$
defining clauses, and so the total number of defining clauses in pog $P$ equals
the number of nonterminal nodes plus the number of edges in the graph, corresponding to our definition of $|P|$.
The number of defining clauses ranged from 304 to 2,761,457,765 with a median of 848,784.

@@

For each of the 124 d-DNNF graphs generated by \dfour{}, we attempted
to use our proof generator and unverified checker, limiting proof
generation to 10,000 seconds.  We also attempted to generate and check
one-sided proofs for the graphs, again with a time limit of 10,000
seconds for proof generation.

@@

Relative to the runtime for \dfour{}, two problems ran faster
with the generator and checker.  These were for problems having small numbers of
models (relative to the number of variables), and so most of the time
spent by both \dfour{} and our proof generator was in running a CDCL
solver to search the very sparse solution space.  
\cadical{} generally outperforms the miniSAT-based solver used by
\dfour{}.  At the other extreme, one problem that required only 19
seconds for \dfour{} required 8657 seconds to generate a proof and 45
seconds to check it.  It is unclear why the proof generation required
so much time.  Overall the ratios between the combined generation plus
checking times versus the time for \dfour{} had a harmonic mean of
5.5.

Of the 16 benchmarks that could not be fully validated, one had so
many defining clauses that it overflowed the 32-bit signed integers
our programs use for clause identifiers.  Another exited due to the
virtual memory limit imposed by the operating system, and the
other 14 timed out.  Of the 15 that did not overflow the clause
counter, we were able to complete a one-sided verification for 9, but
the other 6 timed out.

@@

Figure~\ref{fig:monolithic:clauses} shows the total number of clauses
for the two approaches for the 70 cases where both approaches completed within 1000 seconds.
As can be seen, the monolithic approach consistently produces
shorter proofs, perhaps due to the benefits of the proof trimming by
\dtrim{}.

@@

Overall, it is clear that these two optimizations are worthwhile, and
sometimes critical, to success for some benchmarks.
